Dark matter halo shapes at $z=0$ in the Auriga simulations

### 2019-07-19 Document/paper.tex
edit

\documentclass[a4paper,fleqn,usenatbib]{mnras}
\usepackage[T1]{fontenc}
\usepackage{ae,aecompl}

@@

\newenvironment{Table}
   {\par\bigskip\noindent\minipage{\columnwidth}\centering}
   {\endminipage\par\bigskip}

\title[The shape of dark matter haloes at $z=0$ in Auriga]
{Dark matter halo shape at $z=0$ in the Auriga simulations: radial
  evolution and alignment with the stellar disk}
\author[Jesus Prada,  Jaime E. Forero-Romero, Volker Springel ]{
Jesus Prada,$^{1}$\thanks{E-mail: jd.prada1760@uniandes.edu.co}
Jaime E. Forero-Romero,$^{1}$
Volker Springel$^{2}$
\\
% List of institutions
$^{1}$Departamento de F\'sica, Universidad de los Andes, Cra. 1 No.
18A-10, Edificio Ip, Bogot\'a, Colombia\\
$^{2}$Max-Planck-Institut f\"ur Astrophysik, Karl-Schwarzschild-Str. 1, D-85741 Garching, Germany\\
}

@@

% Abstract of the paper
\begin{abstract}
We present shape measurements of dark matter halos at redshift $z=0$
in a suite of 30 zoom-in simulations from the Auriga project. 
We compare the results in full magnetohydrodynamics against dark
matter only physics. 
We find a strong influence of baryons in making dark matter halos
rounder at all radii compared to its dark matter only counterparts.
At distances $\lesssim 30$ kpc, rounder and coherent dark matter
distributions correlate with extended massive stellar disks and low gas
densities.  
We measure the alignment between the halo and the stellar disk and
find almost perfect alignment at $0.25R_{200}\sim56$ kpc.   
In some cases the alignment between the two components significantly
changes as a function of radius implying that the halo shape twists;
this effect correlates with extended stellar disks and
low gas densities and is almost absent in the dark matter only
simulations.  
In a comparison against observational constraints  we find that $20\%$
of halos are consistent with observational results derived
from the Pal 5 stream that favor an almost spherical shape.
Including baryons is a required element to achieve this
level of agreement. In contrast, none of the simulations (dark matter
only nor with baryons) match the constrains derived from the
Sagittarius stream favoring an oblate dark matter halo.
\end{abstract}

@@

The understanding of our Universe as a whole is greatly influenced by
accurate observations and modeling of our Galaxy. 
Explaining the matter budget and kinematical state of the Milky Way
(MW) is  equivalent to figuring out its formation process in a
cosmological context. 
As a first approximation, the MW morphology and
separation into global kinematically coherent components, such as its
disk and bulge, can be used to support the  existence of a Dark Matter
(DM) component to explain its dynamics around the solar neighborhood
\citep{2000MNRAS.311..361O,2009PASJ...61..227S,2010JCAP...08..004C,2013ApJ...779..115B,Iocco15}.

A more detailed description of the full three-dimensional MW
gravitational potential is possible through the interpretation of 
fossil records of stellar streams.
These remanents, resulting from infalling globular clusters or
satellite galaxies that got tidaly disrutpted by the gravitational
potential of the Milky Way, allow in principle tighter constrains 
on the shape of the dark matter halo in outter regions of our Galaxy
\citep{1998ApJ...495..297J,1999MNRAS.307..495H, 1999MNRAS.307..877T}.

@@

These two questions have been difficult to address because for a long
time it was difficult to produce realistic galactic disks within the
CDM context.
The success of such enterprise has been the result of high numerical
resolution and the understanding that baryonic effects such as stellar
feedback and black hole feedback play an important role in forming a
galactic disk resembling the Milky Way.

Along the way numerical experiments show that the baryonic effects also impact
the dark matter halo shape making the halo rounder than it otherwise
be in a dark matter only simulation
\citep{Dubinski94,Debattista08,Kazantzidis10,Abadi10,Bryan13,Chua19,Artale19}. 
High resolution simulations have also allowed detailed studies of the
alignment between the stellar disk and the dark matter halo
\cite{Bailin05,DeBuhr12,Debattista13,Gomez17}.

Observational constraints of the dark matter halo have been in
tension during the last decade. 
One can find studies favoring prolate
\citep{Banerjee_and_Chanda_2011,Bowden_et_al._2016},  oblate
\citep{LM10,Deg_and_Widrow_2013,Vera-Ciro_and_Helmi_2013} and
spherical configurations \citep{Bovy16}.  
Constraints from modeling of stellar streams discard the prolate
configuration \citep{Pearson_et_al._2015} although some other studies
still doubt that stellar streams can be used to constrain the halo
shape once other assumptions, such as the density profile, are relaxed
\citep{Ibata_et_al._2013}.

@@

\begin{figure}
\begin{center}
\includegraphics[width=0.45\textwidth]{triaxiality_observations.pdf}
\end{center}
\caption{Comparison of our results against other simulations
  by \citet{Chua19} (Illustris) and observational constraints for the 
dark matter halo shape in the Milky Way by \citet{LM10} (LM10) and
\citet{Bovy16} (BBFK16).   
We report our results in the MHD simulations in such a way as to
bracket the radii in the other estimates.
We find that our results are broadly consistent with the Illustris
simulations given the broad dispersion in both data sets.
The consistency with the constraints by \citet{Bovy16} is marginal,
only 1/5 of the halos in our sample seem to be consistent within the
observational reults.
The result by \citet{LM10} would place the Milky Way halo as an
atypical object in the $\Lambda$ CDM context.} 
\label{fig:observations}
\end{figure}

@@

The DM halo shape at a fixed radius is an estimate of either
the isopotential or isodensity surfaces.  
Observational inference models usually estimate the 
isopotential contours which are probed by tracers (gas, stars), while
simulations work with the isodensity contours which can be directly
calculated from particle positions.  
Furthermore, the density contours in thin shells are very sensitive to
the presence of small satelites.  
For this reason we measure the shape by taking
volume-enclosed particles, rather than shell-enclosed.  
This method yields results in good agreement to the isodensity
contours for radii $leq 140$ kpc as explored by
\citep{VeraCiro11}.

In particular, we measure the shape using the reduced inertia tensor
\citep{Allgood_et_al._2006},

@@

Following the convergence criterion by \cite{Vera-Crito_et_al.2011} we
restrict the sampling of the ellipsoidal parameters to radii  between
$\sim 2$kpc and $R_{200}$, where  $R_{200}$ correspond to the  radius
enclosing a sphere with 200 times the critical density of the Universe.
On average, over the 30 halos in the \texttt{Level4} sample
$R_{200}=230\pm 15$kpc. 
For \texttt{Level3} halos we go down to distances of $\sim 0.2$ kpc.

@@

\subsection{Radial trends at $z=0$}

@@

In Figure \ref{fig:triaxial_cumulative} we summarize the results at
different radii using the cumulative distributions for the 
triaxility parameter $T$ defined as 
\begin{equation}
T=\frac{a^2-b^2}{a^2-c^2}.
\label{eq:triaxiality}
\end{equation}
The left panel of this Figure shows that in the DMO sample the
triaxility has a median larger than $0.5$ at all radii, furthermore
this median value increases as we move towards the inner part of the
halo.
The right panel shows the exact complementary picture in the MHD
sampe.
There the median triaxility is alwas smaller than $0.5$ and this
triaxility is smaller as we move closer to the galactic disk.

@@

We now use the results reported by \cite{LM10} and \cite{Bovy16}
to place our results in an observational context.
\cite{LM10} used observations of the Sagittarius tidal stream to
constrain the shape of the gravitational potential.
Their point of depart is that previous studies that assumed an
axisymetric galactic potential were not able to fit all the available
dynamic constraints for the Sagittarius stream, therefore making
necessary the use of a rigid triaxial potential with coaxial potential
ellipsoids for the dark matter component.  
Their results constrain the triaxility of this potential
component. 
They also translate ther results into a triaxiality of the density
contours (that could be compared against our results)
 to be $(c/a)=0.44$ and $(b/a)=0.97$ at a radius of $\sim 40$kpc. 
They do not report any uncertainties for these two values. 
Looking at their plots for the quality of fit criterion as a function
of dark halo axial scales (their Figure 5), we choose a conservative $10\%$
relative uncertaintyy
One surprising element in their results is that  the major axis of the
halo shape is perpendicular to the stellar disk plane.

@@

In this paper we measured the shape of thirty isolated Milky Way like 
dark matter haloes simulated in the Auriga project using the zoom-in
technique. 
The Auriga project simulated these halos with two different setups:
dark matter only (DMO) simulation and full magnetohydrodynamics (MHD)
including star formation and feedback.
We used the shape measurement algorithm by \cite{Allgood06} on the
dark matter halos of these two kinds of simulations to quantify the
halo shape as a function of radius and the degree of aligment between
the stellar disk and the halo.

@@

 \bibliographystyle{mnras}
 \bibliography{references}
\end{document}

>>>>

\documentclass[usenatbib]{mnras}
\usepackage[T1]{fontenc}
\usepackage{ae,aecompl}

@@

\title[Dark matter halo shapes at $z=0$] {Dark matter halo shapes at
  $z=0$ in the Auriga simulations} 
\author[Jesus Prada,  Jaime E. Forero-Romero, Volker Springel ]{
Jesus Prada,$^{1}$\thanks{E-mail: jd.prada1760@uniandes.edu.co}
Jaime E. Forero-Romero,$^{1}$
Volker Springel$^{2}$
\\
% List of institutions
$^{1}$Departamento de F\'isica, Universidad de los Andes, Cra. 1 No.
18A-10, Edificio Ip, Bogot\'a, Colombia\\
$^{2}$Max-Planck-Institut f\"ur Astrophysik, Karl-Schwarzschild-Str. 1, D-85741 Garching, Germany\\
}

@@

% Abstract of the paper
\begin{abstract}
We present shape measurements of Milky Way sized dark matter halos at
redshift $z=0$ in a suite of 30 zoom simulations from the Auriga
project. 
We compare the results in full magnetohydrodynamics against dark
matter only physics. 
We find a strong influence of baryons in making dark matter halos
rounder at all radii compared to its dark matter only counterparts.
At distances $\lesssim 30$ kpc, rounder and coherent dark matter
distributions correlate with extended massive stellar disks and low gas
densities.  
We measure the alignment between the halo and the stellar disk and
find almost perfect alignment at $0.25R_{200}\sim56$ kpc.   
In some cases the alignment between the two components significantly
changes as a function of radius implying that the halo shape twists;
this effect correlates with extended stellar disks and
low gas densities and is almost absent in the dark matter only
simulations.  
In a comparison against observational constraints  we find that $20\%$
of halos are consistent with observational results derived
from the Pal 5 stream that favor an almost spherical shape.
Including baryons is a required element to achieve this
level of agreement. In contrast, none of the simulations (dark matter
only nor with baryons) match the constrains derived from the
Sagittarius stream favoring an oblate dark matter halo.
\end{abstract}

@@

Our physical picture of the Universe as a whole has be shaped by
accurate observations and modeling of our own Galaxy. 
Explaining the matter budget and kinematical state of the Milky Way
(MW) is  equivalent to figuring out its formation process in a
cosmological context. 
As a first approximation, the MW morphology and
separation into global kinematically coherent components, such as its
disk and bulge, can be used to support the  existence of a Dark Matter
(DM) component to explain its dynamics around the solar neighborhood
\citep{2000MNRAS.311..361O,2009PASJ...61..227S,2010JCAP...08..004C,2013ApJ...779..115B,Iocco15}.

A more detailed description of the full three-dimensional MW
gravitational potential is possible through the interpretation of 
fossil records of stellar streams.
These remanents, resulting from infalling globular clusters or
satellite galaxies that got tidaly disrutpted by the gravitational
potential of the Milky Way, allow in principle tighter constrains 
on the shape of the dark matter halo in outter regions of our Galaxy
\citep{1998ApJ...495..297J,1999MNRAS.307..495H, 1999MNRAS.307..877T}
than dynamical measurements close to the solar neighborhodd.

@@

These two questions have been difficult to address because for a long
time it was difficult to produce realistic galactic disks within the
CDM context.
The success of such enterprise has been the result of high numerical
resolution and the understanding that baryonic effects such as stellar
feedback and black hole feedback play an important role in forming a
galactic disk resembling the Milky Way.
Along the way numerical experiments show that the baryonic effects also impact
the dark matter halo shape making the halo rounder than it otherwise
be in a dark matter only simulation
\citep{Dubinski94,Debattista08,Kazantzidis10,Abadi10,Bryan13,Chua19,Artale19}. 
High resolution simulations have also allowed detailed studies of the
alignment between the stellar disk and the dark matter halo
\citep{Bailin05,DeBuhr12,Debattista13,Gomez17}.

Observational constraints of the dark matter halo have been in
tension during the last decade. 
One can find studies favoring prolate
\citep{Banerjee_and_Chanda_2011,Bowden_et_al._2016},  oblate
\citep{LM10,Deg_and_Widrow_2013,Vera-Ciro_and_Helmi_2013} and
spherical configurations \citep{Bovy16}.  
Constraints from modeling of stellar streams discard the prolate
configuration \citep{LM10,Pearson_et_al._2015,Bovy16} although some other studies
still doubt that stellar streams can be used to constrain the halo
shape once other assumptions, such as the density profile, are relaxed
\citep{Ibata_et_al._2013}.

@@

\begin{figure}
\begin{center}
\includegraphics[width=0.45\textwidth]{triaxiality_observations.pdf}
\end{center}
\caption{Comparison of our results against 
observational constraints for the 
dark matter halo shape in the Milky Way by \citet{LM10} (LM10) and
\citet{Bovy16} (BBFK16).   
We find that  1/5 of the halos in the MHD  sample are consistent with
the constraints by \citet{Bovy16}.
In contrast, none of the halos in the MHD and DMO seems to be
consistent with the results by \citet{LM10}.}
\label{fig:observations}
\end{figure}

@@

The DM halo shape at a fixed radius is an estimate of either
the isopotential or isodensity surfaces.  
Observational inference models usually estimate the 
isopotential contours which are probed by tracers (gas, stars), while
simulations work with the isodensity contours which can be directly
calculated from particle positions.  
Furthermore, the density contours in thin shells are very sensitive to
the presence of small satelites.  
For this reason we measure the shape by taking
volume-enclosed particles, rather than shell-enclosed.  
This method yields results in good agreement to the isodensity
contours for radii $\leq 140$ kpc as explored by
\citep{VeraCiro11}.

In particular, we measure the shape using the reduced inertia tensor
\citep{Allgood06},

@@

Following the convergence criterion by \cite{VeraCiro11} we
restrict the sampling of the ellipsoidal parameters to radii  between
$\sim 2$kpc and $R_{200}$, where  $R_{200}$ correspond to the  radius
enclosing a sphere with 200 times the critical density of the Universe.
On average, over the 30 halos in the \texttt{Level4} sample
$R_{200}=230\pm 15$kpc.

@@

\subsection{Radial trends}

@@

In Figure \ref{fig:triaxial_cumulative} we summarize the results at
different radii using the cumulative distributions for the 
triaxility parameter $T$ defined as

\begin{equation}
T=\frac{a^2-b^2}{a^2-c^2}.
\label{eq:triaxiality}
\end{equation}

The left panel of this Figure shows that in the DMO sample the
triaxility has a median larger than $0.5$ at all radii, furthermore
this median value increases as we move towards the inner part of the
halo.
The right panel shows the exact complementary picture in the MHD
sampe.
There the median triaxility is alwas smaller than $0.5$ and this
triaxility is smaller as we move closer to the galactic disk.

@@

We now use the results reported by \cite{LM10} and \cite{Bovy16}
to place our results in an observational context.
\cite{LM10} used observations of the Sagittarius tidal stream to
constrain the shape of the gravitational potential.
Their point of depart is that previous studies that assumed an
axisymetric galactic potential were not able to fit all the available
dynamic constraints for the Sagittarius stream, therefore making
necessary the use of a rigid triaxial potential with coaxial potential
ellipsoids for the dark matter component.  
Their results constrain the triaxility of this potential
component. 
They also translate ther results into a triaxiality of the density
contours (that could be compared against our results)
 to be $(c/a)=0.44$ and $(b/a)=0.97$ at a radius of $\sim 40$kpc. 
They do not report any uncertainties for these two values. 
Looking at their plots for the quality of fit criterion as a function
of dark halo axial scales (their Figure 5), we choose a conservative $10\%$
relative uncertainty.
One surprising element in their results is that  the major axis of the
halo shape is perpendicular to the stellar disk plane.

@@

In this paper we measured the shape of 30 isolated Milky Way like 
dark matter haloes simulated in the Auriga project using the zoom-in
technique. 
The Auriga project simulated these halos with two different setups:
dark matter only (DMO) simulation and full magnetohydrodynamics (MHD)
including star formation and feedback.
We used the shape measurement algorithm by \cite{Allgood06} on the
dark matter halos of these two kinds of simulations to quantify the
halo shape as a function of radius and the degree of aligment between
the stellar disk and the halo.

@@

 \bibliographystyle{mnras}
 \bibliography{references}

\end{document}
